---
title: "Worksheet 3: Grammar of Graphics"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

This semester we will learn how to make different types of visualizations to explore our data. In this worksheet, we will get familiar with principles of \`ggplot, the Grammar of Graphics.

## 1. Dataset and Library

We will consider the built-in dataset `quakes` again. It contains information about earthquakes occurring near Fiji islands since 1964. To refresh our memory, let's take a look:

```{r}
# Take a look at the data frame
head(quakes)
```

We will keep exploring this dataset, using `ggplot` which is also in the `tidyverse` library:

```{r}
# Upload the library
library(tidyverse)
```

We will also use the pipe `|>` which is a very important operator to build on code:

```{r}
# Filter a dataframe
filter(quakes, depth > 300)

# Equivalent to piping quakes into filter()
quakes |> 
  filter(depth > 300)
```

```{r}
# Create a new variable in a dataframe
mutate(quakes, depth_2cat = ifelse(depth > 300, "Deep", "Not deep"))

# Equivalent to piping quakes into mutate()
quakes |> 
  mutate(depth_2cat = ifelse(depth > 300, "Deep", "Not deep"))
```

## 2. Create a `ggplot`

### a. Define a plot

The `ggplot()` function helps us build a graph. Within this function, we specify the dataframe to explore:

```{r}
# Create a ggplot
ggplot(data = quakes)
```

Well, it's empty because we haven't specified how to represent the variables in the `quakes` object...

### b. Add a geometric object

The `geom_...()` functions helps us define how to represent the variables with geometric objects (points, lines, bars, etc.). For example, `geom_point()` can represent our data with points:

```{r, eval = FALSE}
# Create a ggplot with a geometric object
ggplot(data = quakes) +
  geom_point()
```

Well, it does not work because we haven't specified which variables to represent... Using `aes()`, aesthetics, we can define the mapping of the variables to each aesthetic of the plot. We can either define the aesthetics for each geometric object or for the whole plot:

```{r}
# Create a ggplot with mapping variables to a geometric object
ggplot(data = quakes) +
  geom_point(aes(x = long, y = lat))

# which here is equivalent to:
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point()

# Note the importance of placing "+" correctly
```

You can add many layers to the `ggplot` and it is a good practice to put the new layers on a new line (be sure to end a line with +).

```{r}
# Add layers to the plot
ggplot(data = quakes, aes(x = long, y = lat)) + 
  geom_point() + 
  geom_line() 

# Note that we would have to define the mapping for each geom if the mapping is not specified as a whole
ggplot(data = quakes) + 
  geom_point(aes(x = long, y = lat)) +
  geom_line(aes(x = long, y = lat))
```

Other common `geom` functions we will use are:

```{r}
# Make a histogram for numeric variables
ggplot(data = quakes) + 
  geom_histogram(aes(x = mag))

# Make a boxplot for numeric variables
ggplot(data = quakes) + 
  geom_boxplot(aes(x = mag))

# Make a barplot for categorical variables
quakes |> 
  # But first create a categorical variable
  mutate(depth_2cat = ifelse(depth > 300, "Deep", "Not deep")) |> 
  ggplot() + 
  geom_bar(aes(x = depth_2cat))
```

We will learn in the next worksheets in which cases it is more appropriate to use each one.

### d. Map to color, shape, size

We can change the appearance of some characteristics of the geometric object.

```{r}
# What does color vs fill do? 
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point(color = "blue", fill = "orange")
ggplot(data = quakes, aes(x = mag)) +
  geom_histogram(color = "blue", fill = "orange")
```

Other options include `shape`, `size` and `alpha`.

#### **Try it! In the code below, input different values between 0 and 10 for `shape`, `size` and `alpha`. What do these options control?**

```{r}
# What does shape vs size vs alpha do?
ggplot(data = quakes, aes(x = long, y = lat)) +
  geom_point(shape = 0.5, size = 0.5, alpha = 0.5)
```

**Shape changes the shape of the points. Size changes the size. Alpha controls the opacity.**

More importantly, we can change the appearance of some characteristics of the geometric object **depending on the values of some variables**. For example, let's control the size of the points by the magnitude of the earthquake:

```{r}
# Map the size to the mag variable 
ggplot(data = quakes, aes(x = long, y = lat, size = mag)) +
  geom_point()

# Note that the option for size is now within the aes() function
```

#### **Try it! Instead of differentiating the earthquakes based on the size of the magnitude, change the `color` of the points based on depth. Anything you notice in this graph?**

```{r}
# Write and submit code here!
ggplot(data = quakes, aes(x = long, y = lat, color = mag)) +
  geom_point()
```

**Write sentences here.**

Compare your plot to the following plot that considers `depth` as a categorical variable instead:

```{r}
# Use a pipe to apply a function to the object quakes
quakes |> 
  # Pipe into mutate: no need to repeat the name of the dataframe
  mutate(depth_2cat = ifelse(depth > 300, "Deep", "Not deep")) |> 
  # Pipe into ggplot: no need to repeat the name of the dataframe
  ggplot() +
  # Define geom and aesthetics
  geom_point(aes(x = long, y = lat, color = depth_2cat))

# How does the legend of color differ from before?
```

## 3. Layering

### a. Add labels and controlling scales

Plots should be easy to interpret and informative labels are a key element in achieving this goal. The `labs()` function provides customized labels for titles, axes, legends, etc.:

```{r}
# Build a plot...
ggplot(data = quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # ...and add labels
  labs(
    # Title
    title = "Distribution of the depth across the trenches",
    # Subtitle
    subtitle = "For earthquakes that occurred near the Fiji Islands since 1964",
    # Caption with source of data
    caption = "Data obtained from Dr. John Woodhouse, Dept. of Geophysics, Harvard University",
    # Label x-axis and y-axis
    x = "Longitude",
    y = "Latitude",
    # Legend of color
    color = "Depth (km)")
```

### b. Control scales

We can also change the scales (tick marks) for a better readability with functions starting with `scale_`:

```{r}
# Build a plot...
ggplot(data = quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # ...and adjust scales
  scale_x_continuous(limits = c(160,190), breaks = seq(160, 190, 2)) + # enforce limits between 160 and 190, show tick marks every 2 degrees
  scale_y_continuous(limits = c(-40,-10), breaks = seq(-40, -10, 5)) # enforce limits between -40 and -10, show tick marks every 5 degrees
```

### c. Facet

Faceting reproduce a graph for each level of another variable (or combination of variables). Let's take a look at the map for three categories of depths:

```{r}
quakes |> 
  # Create a variable called depth_3cat
  mutate(depth_3cat = case_when(
    depth <= 70 ~ "Shallow",
    depth > 70 & depth <= 300 ~ "Intermediate",
    depth > 300 ~ "Deep")) |> 
  # Create a plot
  ggplot() + 
  geom_point(aes(x = long, y = lat, color = depth_3cat)) +
  # Facet by depth category
  facet_wrap(~depth_3cat)
```

#### **Try it! Compare 3 histograms of magnitude for each of the three categories of depth. Anything you noticed in this graph?**

```{r}
# Write and submit code here!
quakes |> 
  # Create a variable called depth_3cat
  mutate(depth_3cat = case_when(
    depth <= 70 ~ "Shallow",
    depth > 70 & depth <= 300 ~ "Intermediate",
    depth > 300 ~ "Deep")) |> 
  # Create a plot
  ggplot() + 
  geom_histogram(aes(x=mag, color = depth_3cat)) +
  # Facet by depth category
  facet_wrap(~depth_3cat, nrow=3)
```

**Write sentences here.**

### d. Color palettes and themes

We can customize many aspects of our graphs by hand (colors, scales, background color, grid, ...) or we can use some themes or palettes other than the defaults.

To define our own palette of colors for numeric values, we can use `scale_color_gradient()` with setting the color for *low* values and another color for *high* values:

```{r}
# Compare values of magnitude (low: not too dangerous, high: more dangerous)
  ggplot(quakes, aes(x = long, y = lat, color = mag)) + 
  geom_point() +
  # Change the color palette
  scale_color_gradient(low = "yellow", high = "darkred")
```

#### **Try it! Display the depth of earthquakes depending on their location and choosing an appropriate color palette.**

```{r}
# Write and submit code here!

```

**Write sentences here.**

Here is a list of some of the discrete palettes that are color-blind friendly:

```{r}
RColorBrewer::display.brewer.all(colorblindFriendly = TRUE)
```

We can use `scale_color_brewer()` when we are mapping a categorical variable to the aesthetic `color =`:

```{r}
quakes |> 
  # Create a variable called depth_cat
  mutate(depth_3cat = case_when(
    depth <= 70 ~ "Shallow",
    depth > 70 & depth <= 300 ~ "Intermediate",
    depth > 300 ~ "Deep"
  )) |> 
  # Compare categories of depth
  ggplot(aes(x = long, y = lat, color = depth_3cat)) + 
  geom_point() +
  # Change the color palette
  scale_color_brewer(palette = "Set2")
```

### e. Themes

There are so many adjustments we can make to our `ggplot` (change background color, color of axis, color of font, ...) and sometimes it is easier to use some predefined themes:

```{r}
# Still same plot
  ggplot(quakes, aes(x = long, y = lat, color = depth)) + 
  geom_point() +
  # Use the default theme_minimal()
  theme_minimal()
```

Visit this website to find a list of all themes available: <https://ggplot2.tidyverse.org/reference/ggtheme.html>
