---
title: "Worksheet 6: Grammar of Data Wrangling"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we will discuss `dplyr` functions to manipulate our data.

## 1. Dataset and Library

We will consider a new dataset in this worksheet. It is a built-in dataset in the `tidyverse` package so first load the package which also contains the `dplyr` functions:

```{r}
# Upload the package
library(tidyverse)
```

The `txhousing` dataset contains information about the housing market in Texas. Run `?txhousing` in your console for more details and take a look at the dataset:

```{r}
# Take a look at the first few rows of the dataset
head(txhousing) 
```

Get information about the dimensions, types of variables, and some examples of values with `glimpse()`:

```{r}
# Dimensions and structure of the dataset
glimpse(txhousing) 
```

There are `r nrow(txhousing)` rows and `r ncol(txhousing)` in the dataset.

What does one row represent?

**Write sentences here.**

Looking at the documentation with `?txhousing`, we are told that the information about the housing market in Texas was provided by the [TAMU real estate center](https://www.recenter.tamu.edu/). The variables are defined as follows:

| Variables       | Description                                                                                    |
|---------------|---------------------------------------------------------|
| `city`          | Name of multiple listing service (MLS) area                                                    |
| `year`, `month` | Year, Month for the housing market data                                                        |
| `sales`         | Number of sales                                                                                |
| `volume`        | Total value of sales                                                                           |
| `median`        | Median sale price                                                                              |
| `listings`      | Total active listings                                                                          |
| `inventory`     | Amount of time (in months) it would take to sell all current listings at current pace of sales |
| `date`          | Date for the housing market data (`year` + `month` / 12)                                       |

Take a look at general summary statistics with `summary()`:

```{r}
# Try to make sense of the values of the variables in the dataset
summary(txhousing) 
```

Let's manipulate this dataset, with the 6 core `dplyr` functions, including some we already know!

## 2. Operations on rows/observations

Let's consider some `dplyr` functions that apply to the rows/observations of our dataset.

### a. Filter

Use `filter()` to choose rows/observations verifying some conditions:

```{r}
# Filter with one criteria
txhousing |>
  filter(city == "Austin")
```

```{r}
# Filter with multiple criteria
txhousing |>
  filter(city == "Austin", sales <= 1000)

# or using connectors
txhousing |>
  filter(city == "Austin" & sales <= 1000)
```

We can check the number of rows that satisfy the conditions by pipping into `nrow()`:

```{r}
# Filter to check missing values for one variable
txhousing |>
  # Filter with multiple criteria
  filter(city == "Austin", sales <= 1000) |>
  # Count the rows
  nrow()
```

Only 2 rows satisfied these conditions!

### b. Arrange

Use `arrange()` to sort rows/observations for some variables Default is ascending (from least to greatest or alphabetically for categories) but to sort in the other direction use `desc()`.

```{r}
# Sort by number of sales, least-to-greatest 
txhousing |>
  arrange(sales)
```

```{r}
# Sort by number of sales, greatest-to-least (descending order)
txhousing |> 
  arrange(desc(sales))
```

### c. Minimum/Maximum values

Let's try `top_n()` vs `slice_max()`/`slice_min()`, and `top_frac()`.

```{r}
# Select top rows (max values) for a variable
txhousing |> 
  # Use `top_n(number of rows, variables)`
  top_n(n = 4, sales)
```

How does it differ from `slice_max`?

```{r}
# Select top percent of rows (max values) for a variable
txhousing |>
  # Use `slice_max(number of rows, variables)`
  slice_max(n = 4, sales)
```

```{r}
# Select bottom rows (min values) for a variable
txhousing |> 
  # Use `top_n(-number of rows, variables)`
  top_n(n = -4, sales)

# Why did we get more than 4 rows?
```

```{r}
# Select top percent of rows (max values) for a variable
txhousing |>
  # Use `top_frac(proportion of rows, variables)`
  top_frac(n = 0.001, sales)

# Note: for the minimum values add `-` in front of the proportion
```

#### **Try it! When were the lowest 5 numbers of sales for Austin? Display them in chronological order.**

```{r}
# Write and submit code here!

```

**Write sentences here.**

## 3. Operations on columns

Let's consider some `dplyr` functions that apply to the columns/variables of our dataset.

### a. Select

Use `select()` to keep or rename a subset of columns/variables.

```{r}
# Select to keep only some variables
txhousing |>
  # Only see 3 variables
  select(city, year, month, sales)
```

```{r}
# Select to keep columns using indexes of the columns
txhousing |>
  select(1:4)
```

```{r}
# Drop variables using "-"
txhousing |>
  # See all but these 4 variables
  select(-city, -year, -month, -date)
```

```{r}
# Select and rename...
txhousing |>
  # Use `select()` to rename some variables new_name = old_name
  select(Location = city, 
         Calendar_Year = year,
         Month = month,
         Number_of_sales = sales)
```

```{r}
# or just use rename() with the same structure
txhousing |> 
  rename(Location = city, 
         Calendar_Year = year,
         Month = month,
         Number_of_sales = sales)
```

### b. Mutate

Use `mutate()` to create new columns/variables:

```{r}
# Find the mean sale price per row
txhousing |> 
  mutate(mean_price = volume/sales)
```

#### **Try it! Calculate the difference between the average price as calculated above and the median sale price. Are these two measures the same? Why/Why not?**

```{r}
# Write and submit code here!

```

**Write sentences here.**

## 4. Create summaries

Let's consider some `dplyr` functions that can create some summaries for our dataset.

### a. Summarize

Use `summarize()` (or `summarise()` in British!) to calculate summary statistics on columns/variables. Some useful summary functions: `mean()`, `sd()`, `median()`, `IQR()`, `min()`, `max()`, `n()`, `n_distinct()`, `cor()`, ...

```{r}
# Find the mean number of sales
txhousing |>
  summarize(mean_sales = mean(sales, na.rm = T)) # ignore NA values
```

```{r}
# Add more summaries:
txhousing |>
  summarize(
    # the mean
    mean_sales = mean(sales, na.rm = T), 
    # the median
    median_sales = median(sales, na.rm = T),
    # the number of rows
    n_rows = n(),
    # the number of distinct cities in the dataset
    n_cities = n_distinct(city),
    # the correlation between sales and median price
    correlation = cor(sales, median, use = "complete.obs"))
```

#### **Try it! Find the total number of `sales` for Austin in 2009.**

```{r}
# Write and submit code here!

```

**Write sentences here.**

What if we wanted to generate a similar report for each year across all cities in `txhousing`? Let's use a function that allows us to create summaries per subgroup.

### b. Group by

This is one very important function! It enables us to create subgroups and apply a function to all these subgroups For example, find summaries per city and per year:

```{r}
# Find summaries by subgroups 
txhousing |>
  # Each year is a subgroup
  group_by(year) |> 
  # Create summaries for each subgroup
  summarize(total_sales = sum(sales, na.rm = TRUE), # total number of sales
            nb_rows = n()) # count how many rows in each subset 
```

Note that there are less rows in 2015. How could it influence the total number of sales during that year?

**Write sentences here.**

Let's try to be a little more specific and find the total number of sales per year and per month:

```{r}
# Find summaries by subgroups 
txhousing |>
  # Each year/month is a subgroup
  group_by(year, month) |> 
  # Create summaries for each subgroup
  summarize(total_sales = sum(sales, na.rm = TRUE), # total number of sales
            nb_rows = n()) # count how many rows: what does this number correspond to?
```

#### **Try it! Find the total number of `sales` per month across all cities in `txhousing`, but ignoring values from 2015 since there are not complete. Then create a `ggplot` to show how the number of sales may vary per month.**

```{r}
# Write and submit code here!

```

**Write sentences here.**

One more example with a plot! Put comments to describe what each line of code does and add a title+labels:

```{r}
txhousing |> 
  #
  group_by(city) |>
  #
  summarize(total_sales = sum(sales, na.rm = TRUE)) |>
  #
  top_n(n = 10, total_sales) |>
  #
  ggplot(aes(y = reorder(city, total_sales), x = total_sales)) + 
  #
  geom_bar(stat = "identity")
```

*Note that `group_by` is not just use to create summaries but that's its most common use.*
