---
title: "Worksheet 7: Joining"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we will discuss how to join information that can be matched across multiple datasets.

## 1. Datasets and Library

We will discuss joining functions to combine datasets in different ways from the `tidyverse` package:

```{r}
# Load a package
library(tidyverse)
```

Consider the following built-in datasets containing information about some band members of the Beatles and Rolling Stones:

```{r}
# Preview datasets
band_members
band_instruments
band_instruments2
```

## 2. Joining datasets

To join datasets, we first need to identify a **key variable** (a variable, or sometimes a set of variables, that defines a unique row in a dataset). What is the key variable to join the datasets above?

**the name**

### a. Inner join

Join datasets using `inner_join()` to get the information they have in common:

```{r}
# Join 2 datasets with `inner_join()`
inner_join(band_members, band_instruments, by = "name")
```

Why we only get 2 rows?

**Write sentences here.**

### b. Left join

Join datasets using `left_join()` to keep information from the "left" dataset and add information from the "right" dataset:

```{r}
# Join 2 datasets with `left_join()` 
left_join(band_members, band_instruments, by = "name")
```

There was one row in the left dataset that did not appear in the right dataset. How did R handle that?

**Write sentences here.**

#### **Try it! Swap the left and right datasets from above. How do the resulting joined dataset compare?**

```{r}
# Write and submit code here!
left_join(band_instruments, band_members, by = "name")
```

**The order of the left and right datasets matter.**

### c. Right join

This function does the opposite of `left_join()` so it is not widely used.

```{r}
# Join 2 datasets with `right_join()`
right_join(band_members, band_instruments, by = "name")
```

Which `left_join()` function above gave a similar result?

**Write sentences here.**

### d. Full join

Join datasets using `full_join()` to keep information from both datasets:

```{r}
# Join 2 datasets with `full_join()`
full_join(band_members, band_instruments, by = "name")
```

Note how R added missing values for the universities that were in only one of these two datasets.

### e. Anti join

We can use `anti_join()` to get information from the left dataset for which there is no information in the right dataset:

```{r}
# Find missing observations with `anti_join()`
anti_join(band_members, band_instruments, by = "name")
```

Mick did not have an instrument in `band_instruments`.

#### **Try it! Find if any musician did not have a band reported in `band_members`.**

```{r}
# Write and submit code here!

```

**Write sentences here.**

### f. Semi join

What happens if we use `semi_join()`?

```{r}
# Using `semi_join()`
semi_join(band_members, band_instruments, by = "name")
```

The `semi_join()` is like an `inner_join()` but only keeping the variables of the "left" dataset (not very commonly used).

## 3. Considering other options when joining

There are some options and common issues to consider when joining different datasets.

### a. No matching key

Some datasets may refer to the same variable with different names. Consider the following dataset:

What happens if we are joining 2 datasets that have different names for the key variable?

```{r, eval = FALSE}
# Join the two datasets with different key variables
left_join(band_members, band_instruments2, by = "name")
```

There is an error because R did not find any common variable (no key variable). We would need to specify the name of the key in each dataset:

```{r}
# Join the two datasets with different key variables
left_join(band_members, band_instruments2,
          # and specify which variables match across datasets with `c()`
          by = c("name" = "artist"))
```

*Note: The order in which we specify the match for the key variable matters: specify the match from the left dataset to the right dataset. Also, note that only the first name of the key variable is kept in the joined dataset.*

### b. Auto-suffixing

Here is another dataset reporting instruments for each band member:

```{r}
# Consider this new dataset
band_instruments3 <- data.frame(
  name = c("John","Paul","Keith"),
  plays = c("a Steinway upright piano his biggest solo hit Imagine ", 
            "54 different instruments", 
            "10 out of his 3,000 guitars regularly"))
band_instruments3
```

What happens if we are joining 2 datasets with the same variable name that is *not a key* variable?

```{r}
# Join the two variables of instruments played
left_join(band_instruments, band_instruments3, by = "name")
```

Any columns that have the same name in both datasets but are not used to join on will be given suffixes `.x` and `.y` to specify which original dataset they came from (left and right, respectively). You can modify the default suffix:

```{r}
# Join the two variables of instruments played
left_join(band_instruments, band_instruments3, by = "name",
          # To give names to the suffix, use `suffix =`
          suffix = c(".instrument",".fun_fact"))
```

*Note: If the same variable appears in both dataset with the same meaning, it might be a key variable! See section e. below.*

### c. Duplicates

Some datasets may have duplicates for the key variable. Consider the resulting dataset from above:

Here is another dataset reporting instruments for each band member:

```{r}
# Consider this new dataset
band_members2 <- data.frame(
  name = c("Mick","John","Paul","John"),
  band = c("Stones", "Beatles", "Beatles", "Bluesbreakers"))
band_members2
```

#### **Try it! Join the information from `band_members2` to the instruments they play. Is the information contained in the resulting dataset correct?**

```{r}
# Write and submit code here!
left_join(band_members2, band_instruments, by="name")
```

**Not necessarily: be careful with duplicates.**

*Note that it is sometimes useful to add repeating information for some rows that share the same key. We just need to be careful that it makes sense!*

### d. Several key variables

Sometimes one variable is not enough to identify a unique match. Consider these two datasets:

```{r}
# Consider these two new datasets
band_members_last <- band_members2 |>
  mutate(last_name = c("Jagger", "Lennon", "McCartney", "Mayall"))
band_instruments_last <- band_instruments |> 
  mutate(last_name = c("Lennon", "McCartney", "Richards"))

# Take a look
band_members_last
band_instruments_last
```

What key variable(s) should be taken into account to identify a unique row?

**Write sentences here.**

```{r}
# Join 2 datasets with 2 key variables
left_join(band_members_last, band_instruments_last, 
          # List the key variables with `c()`
          by = c("name", "last_name"))
```

------------------------------------------------------------------------

Through this worksheet, we created many datasets to play with, here's one last one!

```{r}
# Information about the bands
band_information <- data.frame(
  main_band = c("Beatles","Stones"),
  plays = c("1960-1970", "1962-present"))
band_information
```

#### **Try it! Join the information from `band_members`, `band_instuments`, `band_information` into one dataset. Does it matter which joining function you use?**

```{r}
# Write and submit code here!
full_join(band_members, band_instruments, by="name") |>
  full_join(band_information, by=c("band"="main_band"), suffix=c(".instrument", ".career"))
```

**Write sentences here.**
