---
title: "Lab 9"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter the names of the group members here:

**This assignment is due by the end of the lab. Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

Let's load the appropriate packages for today:

```{r, warning=FALSE}
library(tidyverse)
library(ggcorrplot)
library(factoextra)
library(cluster)
```

In this lab, we will use a dataset that contains demographic information about 850 customers of a credit card company:

```{r}
Customer_Segmentation <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//Customer_Segmentation.csv")

# Take a quick look!
head(Customer_Segmentation)
```

The `Education` variable is coded as: 1 = high school, 2 = some college, 3 = college degree, 4 = graduate degree. The `Income`, `Card_Debt`, and `Other_Debt` are reported in thousands of dollars.

Suppose that the company is about to advertise new types of credit cards. The goal of the lab is to identify different groups of customers based on the variables available in the dataset.

------------------------------------------------------------------------

### Question 1 (3 pts)

Which variable in the dataset above should we drop in this analysis and why? Create a dataframe called `my_segmentation` without that variable and scale all of the other variables.

**Your answer goes here. Write sentences in bold.**

```{r}
# your code goes below (replace this comment with something meaningful)

```

------------------------------------------------------------------------

### Question 2 (5 pts)

Create a correlation matrix displaying the correlation coefficient for each pair of variables.

```{r}
# your code goes below (replace this comment with something meaningful)

```

Which pair of variables has the strongest positive correlation? Why do you think that makes sense?

**Your answer goes here. Write sentences in bold.**

Which pair of variables has a negative correlation? Why do you think that makes sense?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 3 (12 pts)

We will try to identify groups of customers based on their characteristics in `my_segmentation`. First, find how many clusters you should consider according to the average silhouette width. *Hint: use `fviz_nbclust()`.*

```{r}
# your code goes below (replace this comment with something meaningful)

```

**Your answer goes here. Write sentences in bold.**

Second, use the appropriate number of clusters you found above and apply the clustering algorithm,`kmeans`, on `my_segmentation`:

```{r}
# For reproducible results 
set.seed(322) 

# your code goes below (replace this comment with something meaningful)

```

Third, visualize the clusters with `fviz_cluster`:

```{r}
# your code goes below (replace this comment with something meaningful)

```

Finally, add a variable to `Customer_Segmentation` to assign the appropriate cluster to each customer. Find the mean of each variable in `Customer_Segmentation` for each cluster.

```{r}
# your code goes below (replace this comment with something meaningful)

```

How would you describe an average customer in each cluster?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Question 4 (4 pts)

The credit card company wants to use the clustering results for their marketing campaign advertising for new credit cards (this strategy is called market segmentation and aims at selling more products with less marketing expenses). Here is the advertisement for two new types of credit cards:

-   *CreditStarter* card: "Start Building Your Credit! Our Introductory Credit Card offers a low limit and benefits designed for cautious spenders---begin your credit journey responsibly."

-   *DebtConsolidator* card: "Simplify Your Finances! Transfer balances hassle-free with our Low-Interest Balance Transfer Card, designed to help you consolidate debt and save on interest payments."

As a data scientist at the credit card company, which card would you recommend for the customers in each cluster? Why?

**Your answer goes here. Write sentences in bold.**

------------------------------------------------------------------------

### Formatting: (1 pt)

Make sure the names of all group members are included at the beginning of the document.

Knit your file! You can knit into pdf directly or into html. Once it knits in html, click on `Open in Browser` at the top left of the window pops out. Print your html file into pdf from your browser.

Any issue? Ask other classmates or TA!

Finally, remember to select pages for each question when submitting your pdf to Gradescope and to identify your group members.
