---
title: "Worksheet 9: Strings and Regular expressions"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r global_options, include=FALSE}
# The following code is a basic setup of options for your document
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE, 
                      warning = TRUE,
                      message = FALSE,
                      fig.align = "center", 
                      R.options = list(max.print=50))

# Edit the file starting below

```

In this worksheet, we will focus on manipulating strings and will introduce regular expressions.

## 1. Set up

Let's load `tidyverse` which contains the `stringr` and `lubridate` package:

```{r, message=FALSE}
# Load package 
library(tidyverse)
```

We will refer to some string objects and also manipulate strings within a dataframe, like `txhousing`:

```{r}
# Take a look at the distinct cities in txhousing
txhousing |>
  distinct(city)
```

Let's manipulate strings with functions from the `stringr` package: the name of these functions start with `str_`.

## 2. Strings

Strings are defined with either single quotes `'` or double quotes `"`:

### a. Calculating length

The `str_length()` function can help us find the length of a string:

```{r}
# String length
str_length("abc")

# How is that different?
str_length("a b c")
```

We can apply this function to many strings contained in a vector!

```{r}
# String length of a vector
str_length(txhousing$city)
```

We can also apply this function to create a new variable in a dataframe:

```{r}
# Keep track of the length of a city in a variable
txhousing |>
  # Only see distinct cities
  distinct(city) |>
  # Find the length and create a new variable
  mutate(city_length = str_length(city)) |>
  # Sort from shorter to longer name
  arrange(city_length)
```

### b. Combining strings

We can use `str_c()` to combine two or more strings:

```{r}
# Combine strings
str_c("Happy", "Monday", "!")

# By default, no space but we can add the argument sep = 
str_c("Happy", "Monday", "!", sep = " ")
```

#### **Try it! Add ", TX" to all cities in `txhousing`.**

```{r}
# Write and submit code here!
txhousing |>
  mutate(str_c(city,", TX"))
```

What if we want to combine all the values of one vector/variable together?

```{r}
# Use the argument collapse =
str_c(c("a","b","c"), collapse = "")

# Or separate by a comma and a space
str_c(c("a","b","c"), collapse = ", ")
```

We can get all distinct cities in one object!

```{r}
# Add state information
txhousing |> 
  # Only distinct cities
  distinct(city) |> 
  # Pull the city as a vector
  pull() |> 
  # Collapse all cities together, separated by a comma and a space
  str_c(collapse = ", ")
```

### c. Changing cases

We can change the strings from lower to uppercase and vice-versa (also use sentence case):

```{r}
# To lower case
str_to_lower("Happy Monday!")

# To upper case
str_to_upper("Happy Monday!")

# To sentence case
str_to_sentence("Happy Monday!")
```

Especially useful if there is some inconsistencies in the categories of a variable!

### d. Subsetting strings

We can focus on a subset of a string with `str_sub()` (only works with indexing positions though):

```{r}
# Select a position in the string
str_sub("Happy Monday!", start = 1, end = 5)

# Or count backwards with -
str_sub("Happy Monday!", start = -7, end = -2)
```

We can also split a string by finding a separator:

```{r}
# Split given a pattern
str_split("Happy Monday!", pattern = " ")
```

Note that the resulting object is called a list and is difficult to manipulate within dataframes.

### e. Finding (exact) matches in strings

Let's start finding patterns in strings! We can find if a pattern occurs in our data with `str_detect()`:

```{r}
# Detect the matches
str_detect("Monday", pattern = "day")
```

#### **Try it! Any cities in `txhousing` contain "county" in their name? If you don't find any, try to change the names to lower cases first.**

```{r}
# Write and submit code here!
txhousing |>
  mutate(city=str_to_lower(city)) |>
  filter(str_detect(city, "county")) |>
  distinct(city)
```

**Write sentences here.**

What if we want to replace a pattern with `str_replace()`:

```{r}
# Replace the matches
str_replace("Monday", pattern = "Mon", replacement = "Tues")
```

#### **Try it! Replace `Fort` with `Ft` in the names of the cities.**

```{r}
# Write and submit code here!
txhousing |>
  mutate(city = str_replace(city, pattern="Fort", replacement="Ft")) |>
  filter(str_detect(city, "Ft"))
```

## 3. Regular expressions (Regex)

Regular expressions are used to describe patterns in strings. They're a little weird at first but they can be very useful, especially when we are looking for patterns with some flexibility.

### a. Wildcards

Use `.` to match any character (except a new line):

```{r}
# Detect the matches in a dataframe
txhousing |>
  # Only distinct cities
  distinct(city) |>
  # Filter cities with any character and "ounty"
  filter(str_detect(city, ".ounty"))
```

### b. Anchors

Let's find a match at the beginning of a string with `^` or at the end of a string with `$` :

```{r}
# Detect the matches in a dataframe
txhousing |>
  # Only distinct cities
  distinct(city) |>
  # Filter cities starting with A
  filter(str_detect(city, "^A"))
```

```{r}
# Detect the matches in a dataframe
txhousing |>
  # Only distinct cities
  distinct(city) |>
  # Filter cities ending with n
  filter(str_detect(city, "n$"))
```

#### **Try it! What is the number of cities that start with the initial of your first name?**

```{r}
# Write and submit code here!
txhousing |>
  distinct(city) |>
  filter(str_detect(city, "^C")) |>
  nrow()
```

**Write sentences here.**

### c. Flexible patterns

To look for certain patterns, we will use `[]`. Here are a few useful patterns:

-   `[0-9]` matches any digit

-   `[ ]` matches any single space

-   `[abc]` matches a, b, or c

-   `[a-zA-Z]` matches any letter, lower case or upper case

-   `[a-zA-Z0-9]` matches any alphanumeric character

Let's check any cities with a space in their name:

```{r}
txhousing |>
  # Only distinct cities
  distinct(city) |>
  # Filter cities starting with A or B
  filter(str_detect(city, "^[AB]"))
```

#### **Try it! Find how many city names end with a vowel.**

```{r}
# Write and submit code here!
txhousing |>
  # Only distinct cities
  distinct(city) |>
  # Filter cities starting with A or B
  filter(str_detect(city, "[aeiouAEIOU]$"))
```

**Write sentences here.**

### d. Special characters

In regular expressions, some characters have special meanings (e.g., `.` matches any character, `^` indicates the start of a string, etc.). Sometimes, we may want to search for these special characters themselves rather than their functionality.

To do this, we can "escape" them using a backslash (`\`).

```{r}
# Actually referring to a quote for a string
'\''
```

The trick is that `\` is a special character itself so we sometimes have to use a few of those `\\`:

```{r}
# Compare these two pieces of code:
str_replace_all("Happy Monday.", pattern = ".", replacement = "!")
str_replace_all("Happy Monday.", pattern = "\\.", replacement = "!")
```
